### 2015-10-09 thesis-proposal/problem.tex
proposal revisions

%Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives. %New primitives, and thus new dialects of these languages, should arise very rarely. 
%New primitives (and new \emph{dialects} of the language). %Extending a general-pandurpose language with new primitives, i.e.  forming an extended \emph{dialect} of the language, should, given a sufficiently expressive such language, be exceedingly rare. %Indeed, a diversity of dialects around a language can be taken as evidence against claims of  generality.%, this can be taken as evidence against claims of its generality. %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
%Many languages claim to achieve this ideal. 
%For example, 
%If we measure the generality of a language by how frequently  dialects of the language emerge, 
%all major contemporary languages seem to leave room for improvement. 
%General-purpose programming languages are a dime a. 
%Among the chief goals in the design of a general-purpose programming languages must achieve \emph{stability}: they  very rarely needs to be extended or forked into dialects by the user community surrounding it. Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in. 
Functional programming language designers have long studied small typed lambda calculi to develop a principled understanding of  fundamental metatheoretic issues, like type safety, and to examine the mathematical character of language primitives of interest in isolation. These studies have informed the design of ``full-scale''\footnote{Throughout this work, phrases that should be read as having an intuitive meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} functional languages, which combine several such primitives and also introduce various generalizations and primitive ``embellishments''  motivated by a consideration of human factors. 
For example, major functional languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components are cognitively useful to human programmers. Similarly, these languages build in derived syntactic forms (colloquially, ``syntactic sugar'') that decrease the syntactic cost of  working with common library constructs, like lists.

The hope amongst many language designers is that a limited number of primitives like these will suffice to produce a ``general-purpose'' programming language, i.e. one where programmers can direct their efforts almost exclusively toward the development of libraries for a wide variety of application domains. However, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of ``dialects'' that continue to proliferate around all major contemporary languages. 
In fact, tools that aid in the construction of so-called  ``domain-specific'' language dialects (DSLs)\footnote{In some parts of the literature, such dialects are called ``external DSLs'', to distinguish them from  ``internal'' or ``embedded DSLs'', which are actually  library interfaces that only ``resemble'' distinct dialects \cite{fowler2010domain}.} seem only to be becoming increasingly prominent. 
%\subsection{Why are there so many dialects?}
%{This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today, and instead see a need for new language dialects?

Although some of these dialects extend the semantics of the language that they are based on, the more common situation appears to be that the existing semantic primitives suffice and it is only the \emph{syntactic cost} of expressing one or more constructs of interest using these primitives  that is too high. In response, library providers construct \emph{syntactic dialects} -- dialects that introduce only new derived syntactic forms. 
%Put another way, syntactic dialects can be specified by a context-independent expansion to the existing language that they are based on. 
For example, Ur/Web is a syntactic dialect of Ur (a language that itself descends from ML \cite{conf/pldi/Chlipala10}) that builds in derived forms for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
This is not an isolated example -- there are many other types of data that could similarly benefit from the availability of specialized derived forms (we will consider another example, regular expression patterns, in Sec. \ref{sec:syntax-examples}). 
Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket \cite{Flatt:2012:CLR:2063176.2063195}, which we will discuss in Sec. \ref{sec:syntax-existing-approaches}, have lowered the engineering costs of constructing syntactic dialects in such situations, further contributing to their proliferation.

@@

Dialect designers have instead had to take a less direct approach to have an impact on large-scale software development: they have had to convince the designers in control of comparatively popular languages, like OCaml and Scala, to include some suitable variant of the primitives they've espoused into backwards compatible language revisions. %These decisions are increasingly influenced by community processes, e.g. the Scala Improvement Process.  %This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g. the Scala Improvement Process). 
%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
%Over time, accepting such extensions has caused these languages to balloon in size. 
This \emph{ad hoc} approach is not sustainable, for three main reasons. First,  there are simply too  many potentially useful such primitives, and many of these are only relevant in relatively narrow application domains (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}). Second, primitives introduced earlier in a language's lifespan can end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms. And third, primitives that prove afteer some time to be flawed in some way cannot be removed or replaced without breaking backwards compatibility. %Because there is no data about how useful a construct is in practice until it is available in a major language, decisions about which constructs to include are often informed only by intuition.
%Recalling the words of  Reynolds, which are clearly as relevant today as they were almost half a century ago \cite{Reynolds70}: %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
%\newpage
%\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
%and generality in language design is a serious unsolved 
%problem.}\begin{flushright}--- John Reynolds (1970)\end{flushright}
%\end{quote}

@@

%Verse features a module system taken directly from SML. Unlike SML, the Verse core language is split into a \emph{typed external language} (EL) specified by {type-directed translation} to a minimal \emph{typed internal language} (IL). 
\begin{enumerate}
\item \textbf{Typed syntax macros} (TSMs), introduced in Sec. \ref{sec:tsms}, reduce the need to primitively build in derived concrete syntactic forms specific to library constructs (e.g. list syntax as in SML or XML syntax as in Scala and Ur/Web), by giving library providers static control over the parsing and expansion of delimited segments of textual syntax (at a specified type or parameterized family of types). 
\item \textbf{Type-specific languages} (TSLs), introduced in Sec. \ref{sec:tsls}, further reduce syntactic cost by allowing library providers to associate a TSM with a type declaration and then rely on a local type inference scheme to invoke it implicitly.
%\item \textbf{Metamodules}, introduced in Sec. \ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec. \ref{sec:verse}). %For example, a library provider can implement the type structure of records with a metamodule that:
%\begin{enumerate}
%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and 
%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products). 
%\end{enumerate}
%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
\end{enumerate} 
As vehicles for this work, we plan to formally specify a series of small typed lambda calculi that capture each of the novel primitives that we introduce ``minimally''. For the sake of examples, we will also describe (but not formally specify) a ``full-scale'' functional language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work that we will describe, because Wyvern is a group effort evolving independently in some important ways.} Verse is based semantically on Standard ML, differing mainly in that it uses a local type inference scheme \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala) for reasons that we will return to in Sec. \ref{sec:tsls}. The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of Verse is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs, and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:verse}.

The primitives we propose are forms of \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of terms. Static functions are functions written in a restricted subset of the language (we will discuss the necessary restrictions in Sec. \ref{sec:tsms-limitations}).%Library providers write these static functions using the Verse \emph{static language} (SL).

The design we are proposing also has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language (but did not take a type theoretic approach) \cite{active-libraries-thesis}. %Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design. % static code generation. %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.

@@

\paragraph{Abstract Syntax} The abstract syntax of {patterns}, $r$, over strings, $s$, is specified as below:\[r ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(r; r) ~|~ \textbf{or}(r; r) ~|~ \textbf{star}(r) ~|~ \textbf{group}(r)\]
One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Verse supports these as datatypes:

@@

However, there are some reasons not to expose this representation of patterns directly to clients. First, regular expression patterns are usually identified up to their reduction to a normal form. For example, $\textbf{seq}(\textbf{empty}, r)$ has normal form $r$. It might be useful for patterns with the same normal form to be  indistinguishable from the perspective of client code. Second, it can be useful for performance reasons to maintain additional data alongside patterns (e.g. a corresponding finite automata) without exposing this ``implementation detail'' to clients. Indeed, there may be many ways to represent regular expression patterns, each with different performance trade-offs. For these reasons, a better approach in Verse, as in ML, is to abstract over the choice of representation using  the module system's support for type abstraction. In particular, we can define the following \emph{module signature}, where the type of patterns, \lstinline{t}, is held abstract:
%Notice that it exposes an interface otherwise  to the one available using a case type:

@@

\paragraph{Concrete Syntax} The abstract syntax of patterns is too verbose to be practical  in all but the most trivial examples, so programmers conventionally write patterns using a more concise concrete syntax. For example, the concrete syntax \lstinline{A|T|G|C} corresponds to the following much more verbose pattern expression (assuming some module \lstinline{R : RX} is in scope here and in the remainder of the examples):
\begin{lstlisting}[numbers=none,mathescape=|]
R.Or(R.Str "SSTRAESTR", R.Or(R.Str "SSTRTESTR", R.Or(R.Str "SSTRGESTR", R.Str "SSTRCESTR")))
\end{lstlisting}

>>>>

%Well-designed general-purpose programming languages give library providers the power to  modularly express a wide variety of useful constructs using a comparatively small set of primitives. %New primitives, and thus new dialects of these languages, should arise very rarely. 
%New primitives (and new \emph{dialects} of the language). %Extending a general-pandurpose language with new primitives, i.e.  forming an extended \emph{dialect} of the language, should, given a sufficiently expressive such language, be exceedingly rare. %Indeed, a diversity of dialects around a language can be taken as evidence against claims of  generality.%, this can be taken as evidence against claims of its generality. %Not all useful abstractions can be  realized, when considered comprehensively, as libraries.% The problem of achieving both simplicity and generality simultaneously with a single system remains unsolved. 
%Many languages claim to achieve this ideal. 
%For example, 
%If we measure the generality of a language by how frequently  dialects of the language emerge, 
%all major contemporary languages seem to leave room for improvement. 
%General-purpose programming languages are a dime a. 
%Among the chief goals in the design of a general-purpose programming languages must achieve \emph{stability}: they  very rarely needs to be extended or forked into dialects by the user community surrounding it. Instead, its users are able to modularly  express a wide variety of useful constructs in libraries, using the comparatively small set of primitives that the language builds in. 
Functional programming language designers have long studied small typed lambda calculi to develop a principled understanding of language-theoretic issues, like type safety, and to examine the mathematical character of language primitives of possible interest in isolation. These studies have then informed the design of ``full-scale''\footnote{Throughout this work, words and phrases that should be read as having an intuitive or informal meaning, rather than a strict mathematical meaning, will be introduced with quotation marks.} functional languages, which combine several such primitives and also introduce various generalizations and primitive ``embellishments''  motivated by a consideration of various human factors. 
For example, major functional languages like Standard ML (SML) \cite{mthm97-for-dart,harper1997programming}, OCaml \cite{ocaml-manual} and Haskell \cite{jones2003haskell} all primitively build in record types, generalizing the nullary and binary product types that suffice in simpler calculi, because explicitly labeled components are cognitively useful to human programmers. Similarly, these languages build in derived syntactic forms (colloquially, ``syntactic sugar'') that decrease the syntactic cost of working with common library constructs, like lists.

The hope amongst many language designers is that a limited number of primitives like these will suffice to produce a ``general-purpose'' programming language, i.e. one where programmers can direct their efforts almost exclusively toward the development of libraries for a wide variety of application domains. However, a stable language design that fully achieves this ideal has yet to emerge, as evidenced by the diverse array of ``dialect'' languages that continue to proliferate around all major contemporary languages. 
In fact, tools that aid in the construction of so-called  ``domain-specific'' language dialects (DSLs)\footnote{In some parts of the literature, such dialects are called ``external DSLs'', to distinguish them from  ``internal'' or ``embedded DSLs'', which are actually  library interfaces that only ``resemble'' distinct dialects \cite{fowler2010domain}.} seem only to be becoming increasingly prominent. 
%\subsection{Why are there so many dialects?}
%{This calls for an investigation}: why is it that programmers and researchers are still so often unable to satisfyingly express the constructs that they need in libraries, as modes of use of the ``general-purpose'' primitives already available in major languages today, and instead see a need for new language dialects?

Some of these dialects extend the type structure of the language that they are based on. However, the more common situation, and the one that we will be our focus in this work, appears to be that the existing semantic primitives suffice and it is only the \emph{syntactic cost} of expressing one or more constructs of interest using these primitives  that is too high. In response, library providers construct \emph{syntactic dialects} -- dialects that introduce only new derived syntactic forms. 
%Put another way, syntactic dialects can be specified by a context-independent expansion to the existing language that they are based on. 
For example, Ur/Web is a syntactic dialect of Ur (a language that itself descends from ML \cite{conf/pldi/Chlipala10}) that builds in derived forms for SQL queries, HTML elements and other datatypes used in the domain of web programming \cite{conf/popl/Chlipala15}. %Syntactic cost is often assessed qualitatively \cite{green1996usability}, though quantitative metrics can be defined. 
This is not an isolated example -- there are many other types of data that could similarly benefit from the availability of specialized derived forms (we will consider another example, regular expression patterns, in Sec. \ref{sec:syntax-examples}). 
Tools like Camlp4 \cite{ocaml-manual}, Sugar* \cite{erdweg2011sugarj,erdweg2013framework} and Racket \cite{Flatt:2012:CLR:2063176.2063195}, which we will discuss in Sec. \ref{sec:syntax-existing-approaches}, have lowered the engineering costs of constructing syntactic dialects in such situations, further contributing to their proliferation.

@@

Dialect designers have instead had to take a less direct approach to have an impact on large-scale software development: they have had to convince the designers in control of comparatively popular languages, like OCaml and Scala, to include some suitable variant of the primitives they've espoused into backwards compatible language revisions. %These decisions are increasingly influenced by community processes, e.g. the Scala Improvement Process.  %This approach concentrates power as well as responsibility over maintaining metatheoretic guarantees in the hands of a small group of language designers, though increasingly influenced by various community processes (e.g. the Scala Improvement Process). 
%Dialects thus serve the role of rhetorical vehicles for new ideas, rather than direct artifacts. 
%Over time, accepting such extensions has caused these languages to balloon in size. 
This \emph{ad hoc} approach is not sustainable, for three main reasons. First,  there are simply too  many potentially useful such primitives, and many of these are only relevant in relatively narrow application domains (for derived syntax, our group has  gathered initial data speaking to this \cite{TSLs}). Second, primitives introduced earlier in a language's lifespan can end up monopolizing finite ``syntactic resources'', forcing subsequent primitives to use ever more esoteric forms. And third, primitives that prove after some time to be flawed in some way cannot be removed or replaced without breaking backwards compatibility. %Because there is no data about how useful a construct is in practice until it is available in a major language, decisions about which constructs to include are often informed only by intuition.
%Recalling the words of  Reynolds, which are clearly as relevant today as they were almost half a century ago \cite{Reynolds70}: %This approach is antithetical to the ideal of a truly \emph{general-purpose language} described at the beginning of this section.
%\newpage
%\begin{quote}\textit{The recent development of programming languages suggests that the simul\-taneous achievement of simplicity 
%and generality in language design is a serious unsolved 
%problem.}\begin{flushright}--- John Reynolds (1970)\end{flushright}
%\end{quote}

@@

%Verse features a module system taken directly from SML. Unlike SML, the Verse core language is split into a \emph{typed external language} (EL) specified by {type-directed translation} to a minimal \emph{typed internal language} (IL). 
\begin{enumerate}
\item \textbf{Typed syntax macros} (TSMs), introduced in Sec. \ref{sec:tsms}, reduce the need to primitively build in derived concrete syntactic forms specific to library constructs (e.g. list syntax as in SML or XML syntax as in Scala and Ur/Web), by giving library providers static control over the parsing and expansion of delimited segments of textual syntax (at a specified type or parameterized family of types). 
\item \textbf{Type-specific languages} (TSLs), introduced in Sec. \ref{sec:tsls}, further reduce syntactic cost by allowing library providers to associate a TSM with a type declaration and then rely on a local type inference scheme to invoke that TSM implicitly.
%\item \textbf{Metamodules}, introduced in Sec. \ref{sec:metamodules}, reduce the need to primitively build in the type structure of constructs like records (and variants thereof),  labeled sums and other interesting constructs that we will introduce later by giving library providers programmatic ``hooks'' directly into the semantics, which are specified as a \emph{type-directed translation semantics} targeting a small \emph{typed internal language} (introduced in Sec. \ref{sec:verse}). %For example, a library provider can implement the type structure of records with a metamodule that:
%\begin{enumerate}
%\item introduces a type constructor, \lstinline{record}, parameterized by finite mappings from labels to types, and defines, programmatically, a translation to unary and binary product types (which are built in to the internal language); and 
%\item introduces operators used to work with records, minimally record introduction and elimination (but perhaps also various functional update operators), and directly implements the logic governing their typechecking and translation to the IL (which builds in only nullary and binary products). 
%\end{enumerate}
%We will see direct analogies between ML-style modules (which our mechanisms also support) and metamodules later.
\end{enumerate} 
As vehicles for this work, we plan to formally specify a series of small typed lambda calculi that capture each of the novel primitives that we introduce ``minimally''. For the sake of examples, we will also describe (but not formally specify) a ``full-scale'' functional language called Verse.\footnote{We distinguish Verse from Wyvern, which is the language referred to in prior publications about some of the work that we will describe, because Wyvern is a group effort evolving independently in some important ways.} Verse is based semantically on Standard ML, differing mainly in that it uses a local type inference scheme \cite{Pierce:2000:LTI:345099.345100} (like, for example, Scala) for reasons that we will return to in Sec. \ref{sec:tsls}. The reason we will not follow Standard ML \cite{mthm97-for-dart} in giving a complete formal specification of Verse is both to emphasize that the primitives we introduce can be considered for inclusion in a variety of language designs, and to avoid distracting the reader with specifications for ``orthogonal'' primitives that are already well-understood in the literature. %We anticipate that future full-scale language specifications will be able to combine the ideas  in the proposed work without trouble. %The purpose of the work being proposed is to serve as a reference for those interested in the new constructs we introduce, not to serve as a language specification. 
%We will give a brief overview of these languages are organized in Sec. \ref{sec:verse}.

The primitives we propose perform \emph{static code generation} (also sometimes called \emph{static} or \emph{compile-time metaprogramming}), meaning that the relevant rules in the static semantics of the language call for the evaluation of \emph{static functions} that generate static representations of terms. Static functions are functions written in a restricted subset of the language (we will discuss the design space of restrictions in Sec. \ref{sec:tsms-limitations}). %Library providers write these static functions using the Verse \emph{static language} (SL).  
The design we are proposing also has conceptual roots in earlier work on \emph{active libraries}, which similarly envisioned using compile-time computation to give library providers more control over aspects of the language (but did not take a type theoretic approach) \cite{active-libraries-thesis}. %Maintaining a separation between the static (or ``compile-time'') phase and the dynamic (or ``run-time'') phase is an important facet of Verse's design. % static code generation. %We will  also introduce a simple variant of each of these primitives that leverages Verse's support for local type inference to further reduce syntactic cost in certain common situations.

@@

\paragraph{Abstract Syntax} The abstract syntax of {patterns}, $r$, over strings, $s$, is specified below:\[r ::= \textbf{empty} ~|~ \textbf{str}(s) ~|~ \textbf{seq}(r; r) ~|~ \textbf{or}(r; r) ~|~ \textbf{star}(r) ~|~ \textbf{group}(r)\]
One way to express this abstract syntax is by defining a recursive sum type \cite{pfpl}. Verse supports these as datatypes:

@@

However, there are some reasons not to expose this representation of patterns directly to clients. First, regular expression patterns are usually identified up to their reduction to a normal form. For example, $\textbf{seq}(\textbf{empty}, r)$ has normal form $r$. It might be useful for patterns with the same normal form to be  indistinguishable from the perspective of client code. Second, it can be useful for performance reasons to maintain additional data alongside regexes (e.g. a corresponding finite automata) without exposing this ``implementation detail'' to clients. Indeed, there may be many ways to represent regular expression patterns, each with different performance trade-offs. For these reasons, a better approach in Verse, as in ML, is to abstract over the choice of representation using  the module system's support for type abstraction. In particular, we can define the following \emph{module signature}, where the type of patterns, \lstinline{t}, is held abstract:
%Notice that it exposes an interface otherwise  to the one available using a case type:

@@

\paragraph{Concrete Syntax} The abstract syntax of patterns is too verbose to be practical  in all but the most trivial examples, so programmers conventionally write patterns using a more concise concrete syntax. For example, the concrete syntax \lstinline{A|T|G|C} corresponds to the following much more verbose pattern expression (assuming some module \lstinline{R : RX} is in scope here and in the remainder of the document):
\begin{lstlisting}[numbers=none,mathescape=|]
R.Or(R.Str "SSTRAESTR", R.Or(R.Str "SSTRTESTR", R.Or(R.Str "SSTRGESTR", R.Str "SSTRCESTR")))
\end{lstlisting}
